---  
title: "Assignment 1"
author: Michael Raminski  
date: "October 11, 2017"  
output: 
  html_document:  
    toc: yes  

---  
email address: mraminski@gmail.com
  
***
Instructions: answer the questions below,  Do not use the R interpreter or the Internet during the quiz  

Use only base R code in your answers , no use of packages unless specified.  No use of for (or any other type of) loops on any questions.

(additional instructions and/or comments)


# Problem 1  
  
Given a 10 element numeric vector v, give an R statement that adds the mean of v to every individual element of v and assigns the result to v1  

```{r, comment=NA}
    v <- c(1:10)
    v1 <- v + mean(v)
    v1
```
  
# Problem 2  
  
Given a character vector vchars, write a single statement that prints the number of missing values in vchar  
  
```{r, comment=NA}
    vchars <- c("a", "b", "c", NA, "d", "e", NA)
    missingvchars <- sum(is.na(vchars))
    missingvchars
```

# Problem 3  
  
Given a numeric vector nvec, write a single R statement that assigns a zero to every negative element present in nvec
  
```{r, comment=NA}
    nvec <- c(-5:5)
    nvec[nvec < 0] <- 0
    nvec
```

# Problem 4  
  
Given three vectors a, b, and d, create a fourth vector consisting of the concatenation of d, b, and a in that order
  
```{r, comment=NA}
    a <- c(3:1)
    b <- c(6:4)
    d <- c(9:7)
    c(d,b,a)
```

# Problem 5  
  
What is printed by the following R statement?
print(typeof(c(TRUE, 317, 412, "2007-03-01"))

```{r, comment=NA}
    print(typeof(c(TRUE, 317, 412, "2007-03-01")))
```

# Problem 6
  
A data frame df has the following columns:

(A) Write a single R statement that deletes all the rows in df for which scores is less than the mean of scores

```{r, comment=NA}
    scores <- c(79.51508, 67.84057, 74.89246, 75.71270, 74.18168, 75.73072, 77.76586, 76.790, 78.888888, 79.999999)
    eval <- c("k", "p", "i", "d", "c", "z", "m", "j", "t", "u")
    df <- data.frame(scores, eval)
    df
    df[df$scores > mean(df$scores),]
```

(B) Add a new column names new_scores to df consisting of a normally distributed random vector with mean 100 and standard deviation 5

```{r, comment=NA}
    df$new_scores <- rnorm(10,100,5)
    df
```

# Problem 7  
  
What is the R statement used to install a package?  Load a package?  Should the package be installed every time it is loaded?

```{r, comment=NA, warning=FALSE}
    #install.packages("dplyr")
    library(dplyr)
    #the package does not need to be re-installed each time it is loaded
```

# Problem 8
  
Write an R function normalize that normalizes a numeric vector or matrix between 0 and 1 using the following formula:

x - min(x) / max(x) - min(x)

Use normalize to normalize a vector v with 10 uniformly distributed random numbers between 100 and 200

```{r, comment=NA}
    normalize_0_1 <- function(x)
    (x-min(x)) / (max(x) - min(x))
    v <- runif(10,100,200)
    normalize_0_1(v) 
```
